---
title: "Untitled"
author: "Amy Gill"
date: "8/8/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Transcript quantification assessment

```{r}
# exon length
lengths = c(100,200,300,100,100)

# exon pattern in transcripts 1-3
mat = cbind(c(1,1,0,1,0),c(1,1,1,1,1),c(0,1,1,0,1))

# length of transcripts
lengths %*% mat

# 1000 reads
w = 1000
 
# counts per exon
counts = c(125,350,300,125,100)

# guess at theta for maximum likelihood estimate
theta.hat = c(1, 2, 3) / 10000

# result of guess - counts too low and out of proportion
mat %*% theta.hat * lengths * w

# roughly estimate theta by solving system of equations
LHS = counts/(lengths * w)
lm.fit(mat, LHS)$coefficients
```

### Q1

```{r}
theta.hat = c(.00075, .0005, .0005)
mat %*% theta.hat * lengths * w
```

### Q2

```{r}
counts = c(60,320,420,60,140)
LHS = counts/(lengths * w)
lm.fit(mat, LHS)$coefficients
```

## Transcriptome alignment assessment

```{r}
genes <- read.table("SRR1039508.genes.results", header=TRUE)
isoforms <- read.table("SRR1039508.isoforms.results", header=TRUE)
```

```{r}
fpkm.per.gene <- split(isoforms$FPKM, isoforms$gene_id)
head(sapply(fpkm.per.gene, sum))
head(genes$FPKM)
```

```{r}
library(dplyr)
isoforms %>% group_by(gene_id) %>% summarize(sum = sum(FPKM))
```

```{r}
genes2 <- genes[genes$FPKM > 0,]
genes2$gene_id <- droplevels(genes2$gene_id)
isoforms2 <- isoforms[isoforms$gene_id %in% genes2$gene_id,]
isoforms2$gene_id <- droplevels(isoforms2$gene_id)
```

```{r}
stopifnot(all(genes2$gene_id == levels(isoforms2$gene_id)))
```

### Q1

```{r}
median(genes2$FPKM)
```

### Q2

```{r}
max.iso = sapply(split(isoforms2$IsoPct, isoforms2$gene_id), max)
mean(max.iso > 95)

# ...or with dplyr:
isoforms2 %>% group_by(gene_id) %>% summarize(max.iso = max(IsoPct)) %>% summarize(answer = mean(max.iso > 95))
```

### Q3

```{r}
boxplot(split(log10(genes2$FPKM), cut(max.iso, 5)), xlab="max.iso", ylab="log10 FPKM")
```

### Q4

```{r}
num.iso <- as.numeric(table(isoforms2$gene_id))
plot(num.iso, max.iso)
```

```{r}
barplot(table(num.iso))
barplot(table(num.iso[num.iso < 15]))
```

## Install bioc assessment

```{r}
BiocManager::version()
```

```{r}
BiocManager::install()
```

## Count matrix assessment

```{r}
library(pasillaBamSubset)
bam.file <- untreated3_chr4()
library(Rsamtools)
bf <- BamFile(bam.file)
library(TxDb.Dmelanogaster.UCSC.dm3.ensGene)
txdb <- TxDb.Dmelanogaster.UCSC.dm3.ensGene
```

### Q1

```{r}
ebg = exonsBy(txdb, by="gene")
seqnames(ebg[1])
```

### Q2

```{r}
chr4.idx <- all(seqnames(ebg) == "chr4")
ebg.sub <- ebg[chr4.idx]
```

```{r}
library(GenomicAlignments)
se = summarizeOverlaps(ebg.sub, bf, ignore.strand=TRUE, singleEnd=FALSE)
head(assay(se), 1)
```

## Normalization and transformation assessment

### Q1

```{r}
download.file("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/wang_eset.RData", "wang_eset.RData")
```

```{r}
load("wang_eset.RData")
library(Biobase)
count.matrix <- exprs(wang.eset)[,10:21]
col.data <- pData(wang.eset)[10:21,]
library(DESeq2)
dds <- DESeqDataSetFromMatrix(count.matrix, col.data, design=~cell.type)
```

```{r}
dds$cell.type
```

```{r}
dds = estimateSizeFactors(dds)
dds$cell.type[which.max(sizeFactors(dds))]
```

### Q2

```{r}
vsd = varianceStabilizingTransformation(dds, blind=FALSE)
plotPCA(vsd, intgroup="cell.type")
```

### Q3

```{r}
rmeans <- rowMeans(assay(vsd)) # row mean of rlog-transformed data
idx <- c(1,2,10,7,8,9,12) # pick some samples for visualization
mat <- assay(vsd)[rmeans > 1,idx] # pull out a small matrix of rlog-transformed counts
colnames(mat) <- vsd$cell.type[idx] # name the columns of matrix by cell type
```

```{r}
panel.sub <- function(x,y,...) points(cbind(x,y)[sample(length(x),1000),],...)
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)  {
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r)
}
pairs(mat, asp=1, col=rgb(0,0,0,.3), lower.panel=panel.cor, upper.panel=panel.sub)
```

```{r}
colnames(mat[,-7])[ which.max(cor(mat[,7], mat[,-7])) ]
```

## DE assessment 1

```{r}
load("wang_eset.RData")
library(Biobase)

count.matrix <- exprs(wang.eset)[,10:21]
col.data <- pData(wang.eset)[10:21,]
library(DESeq2)
dds <- DESeqDataSetFromMatrix(count.matrix, col.data, design=~cell.type)
```

```{r}
dds$type <- factor(ifelse(dds$cell.type %in% c("cerebellum","mixed.brain"), 
                          "brain", "other"))
dds$type <- relevel(dds$type, "other")

design(dds) <- ~ type
```

### Q1

```{r}
dds = DESeq(dds)
res = results(dds)
rownames(res)[which.min(res$padj)]
```

### Q2

```{r}
res2 = results(dds, lfcThreshold=2)
summary(res2)
```

```{r}
table(res2$padj < 0.1 & res2$log2FoldChange > 0)
```

## DE assessment 2

```{r}
plotCounts(dds, which.min(res$padj), intgroup="type")
```

```{r}
par(mfrow=c(3,3))
for (i in 1:9)  plotCounts(dds, order(res$padj)[i], intgroup="type")
```

```{r}
par(mfrow=c(1,1))
```

```{r}
top <- rownames(res)[head(order(res$stat, decreasing=TRUE), 20)]
```


### Q1

```{r}
library(org.Hs.eg.db)                              
select(org.Hs.eg.db, keys=top, columns="SYMBOL", keytype="ENSEMBL")
```

### Q2

```{r}
select(org.Hs.eg.db, keys=top, columns="GENENAME", keytype="ENSEMBL")
```

## SVA assessment

```{r}
download.file("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bottomly_eset.RData", "bottomly_eset.RData")
```

```{r}
load("bottomly_eset.RData")
```


```{r}
count.matrix <- exprs(bottomly.eset)
col.data <- pData(bottomly.eset) 
library(DESeq2)
dds <- DESeqDataSetFromMatrix(count.matrix, col.data, design=~strain)
```

```{r}
dds$num.tech.reps <- factor(dds$num.tech.reps)
```

```{r}
library(sva)
dat <- counts(dds, normalized=TRUE)
idx <- rowMeans(dat) > 1
dat <- dat[idx,]
mod <- model.matrix(~ strain, colData(dds))
mod0 <- model.matrix(~ 1, colData(dds))
svseq <- svaseq(dat, mod, mod0, n.sv=2)

```

## Exon usage assessment

```{r}
library("pasilla")
inDir = system.file("extdata", package="pasilla", mustWork=TRUE)        
countFiles = list.files(inDir, pattern="fb.txt$", full.names=TRUE)    
flattenedFile = list.files(inDir, pattern="gff$", full.names=TRUE) 

sampleTable = data.frame(row.names = c( "treated1", "treated2", "treated3","untreated1", "untreated2", "untreated3", "untreated4" ), condition = c("knockdown", "knockdown", "knockdown", "control", "control", "control", "control" ))                 

library("DEXSeq")  
dxd = DEXSeqDataSetFromHTSeq(countFiles, sampleData=sampleTable, 
  design= ~ sample + exon + condition:exon, flattenedfile=flattenedFile )

rmean = rowMeans(counts(dxd))  
dxd2L = dxd[seqnames(rowRanges(dxd)) == "chr2L" & rmean > 10,] 

dxd2L = dxd2L[1:1000,]  
```

### Q1

```{r}
dxd2L = estimateSizeFactors( dxd2L )                                                                
dxd2L = estimateDispersions( dxd2L )                                                                
dxd2L = testForDEU( dxd2L )                                                                         
dxd2L = estimateExonFoldChanges( dxd2L, fitExpToVar="condition")                                    
dxr = DEXSeqResults( dxd2L )                                                                        
rownames(dxr)[ which.min(dxr$padj) ]
```

## Isoform abundance assessment

```{r}
library(cummeRbund)
myDir <- system.file("extdata", package="cummeRbund") 
gtfFile <- system.file("extdata/chr1_snippet.gtf",package="cummeRbund")
cuff <- readCufflinks(dir=myDir,gtfFile=gtfFile,genome="hg19",rebuild=TRUE)
gene.features <- annotation(genes(cuff))
head(gene.features)
isoforms.per.gene <- gene.features[!duplicated(gene.features$isoform_id),    c("gene_id","isoform_id")]
isoforms.per.gene <- isoforms.per.gene[order(isoforms.per.gene$isoform_id),]
head(isoforms.per.gene)
```

```{r}
gene.tab <- table(isoforms.per.gene$gene_id)
```

### Q1

```{r}
table(gene.tab)
```

### Q2

```{r}
isoform.fpkm <- fpkm(isoforms(cuff))
head(isoform.fpkm)
table(isoform.fpkm$sample_name)
```

```{r}
ips <- isoform.fpkm[isoform.fpkm$sample_name == "iPS",]
hesc <- isoform.fpkm[isoform.fpkm$sample_name == "hESC",]
stopifnot(all(ips$isoform_id == isoforms.per.gene$isoform_id))
stopifnot(all(hesc$isoform_id == isoforms.per.gene$isoform_id))
```

### Q2

```{r}
ips.max = sapply(split(ips$fpkm, isoforms.per.gene$gene_id), which.max)    
hesc.max = sapply(split(hesc$fpkm, isoforms.per.gene$gene_id), which.max)
mean(ips.max == hesc.max) 
```

### Q3

```{r}
ips.max = sapply(split(ips$fpkm, isoforms.per.gene$gene_id), which.max)    
hesc.max = sapply(split(hesc$fpkm, isoforms.per.gene$gene_id), which.max)
mean((ips.max == hesc.max)[ gene.tab > 1 ]) 
```

## DNA Methylation assessment

### Q1

```{r}
BiocManager::version()
```

### Q2

```{r}
BiocManager::install("minfi")
package.version("minfi")
```

### Q5

```{r}
library(BSgenome.Hsapiens.UCSC.hg19)
chr22 = Hsapiens[["chr22"]]
s = subseq(chr22,start=23456789,width=1000)
print( as.character(s) )
```
```{r}
letterFrequency(s, "CG", as.prob=TRUE)
```

### Q6

```{r}
countPattern("CG",s)
```

### Q7

```{r}
countPattern("GC",s)
```

## CpG island assessment

```{r}
library(AnnotationHub)
ah = AnnotationHub()
head(ah)
```

```{r}
ah = subset(ah,ah$genome=="hg19")
```

```{r}
query(ah,"genes")
```

### Q1

```{r}
query(ah, "cpg Islands")
```

### Q2

```{r}
library(AnnotationHub)
ah = AnnotationHub()
cgi = ah[["AH5086"]]
```

```{r}
class(cgi)
```

### Q3

```{r}
length(cgi)
```

### Q4

```{r}
library(BSgenome.Hsapiens.UCSC.hg19)
cgiseq= getSeq(Hsapiens,cgi)
```

```{r}
genome(cgi)[1:24]
genome(Hsapiens)[1:24]
```

```{r}
res = alphabetFrequency(cgiseq)
L = rowSums(res)
cprop = res[,"C"]/L
median(cprop)
# or
median(letterFrequency(cgiseq, "C", as.prob=TRUE))
# Make a histogram to see they are generally high
hist(cprop)
```

### Q5

```{r}
res = alphabetFrequency(cgiseq)
L = rowSums(res)
gprop = res[,"G"]/L
median(gprop)
##Make a histogram to see they are generally high
hist(gprop)
```

### Q6

```{r}
library(BSgenome.Hsapiens.UCSC.hg19)
cgiseq= getSeq(Hsapiens,cgi)

res = alphabetFrequency(cgiseq)
L = rowSums(res)
cprop = res[,"C"]/L
gprop = res[,"G"]/L
expected = L * cprop * gprop
observed = vcountPattern("CG",cgiseq)
cpgoe=observed/expected
median(cpgoe)
## We can look at a histogram
hist(cpgoe)
abline(v=1,col=2)
##because these are ratios, better to look at log
hist(log2 ( cpgoe ))
abline(v=0,col=2)
```

```{r}
observed = vcountPattern("GC",cgiseq)
gpcoe=observed/expected
median(gpcoe)
### We can look at a histogram
boxplot(list(cpgoe,gpcoe))
```

### Q8

```{r}
chr2use = seqlevels(cgi)[1:24]
index = which( seqnames(cgi) %in% chr2use)
noncgi = shift(cgi[index],20000)
library(BSgenome.Hsapiens.UCSC.hg19)
noncgiseq= getSeq(Hsapiens,noncgi)

nullres = alphabetFrequency(noncgiseq)
keepIndex=nullres[,"G"]>0 &  nullres[,"C"]>0 & nullres[,"N"]==0
nullres = nullres[keepIndex,]
noncgiseq=noncgiseq[keepIndex]
```

```{r}
L2 = rowSums(nullres)
cprop2 = nullres[,"C"]/L2
gprop2 = nullres[,"G"]/L2
expected2 = L2 * cprop2 * gprop2
observed2 = vcountPattern("CG",noncgiseq)
noncgioe=observed2/expected2
median(noncgioe)
## We can compare them
boxplot(gpcoe,noncgioe,cpgoe)
```

## Finding differentially methylated regions in R assessment

```{r}
library(devtools)
install_github("genomicsclass/coloncancermeth")
```

```{r}
library(coloncancermeth)
data(coloncancermeth)
dim(meth)
dim(pd)
print( gr )
```

### Q1

```{r}
table(pd$Status)
```

```{r}
names(pd)[ grep("barcode",names(pd)) ]
##
which(pd[,1]=="TCGA-A6-4107" & pd$Status=="cancer")
```


### Q2

```{r}
d = dist( t(meth))
mds = cmdscale(d)
cols = as.numeric(pd$Status)
plot(mds,col=cols)
legend("topleft",levels(pd$Status),col=1:2,pch=1)
```

### Q3

```{r}
library(limma)
X = model.matrix(~pd$Status)
fit = lmFit(meth,X)
eb = eBayes(fit)
pvals = eb$p.value[,2]
```


```{r}
library(qvalue)
qvals = qvalue(pvals)$qvalue
mean(qvals<=0.05)
```

### Q4

```{r}
library(qvalue)
library(limma)
X = model.matrix(~pd$Status)
fit = lmFit(meth,X)
eb = eBayes(fit)
pvals = eb$p.value[,2]
qvals = qvalue(pvals)$qvalue

index = which(qvals <=0.05)
diffs = fit$coef[index,2]
mean(diffs > 0)
```

### Q5

```{r}
library(AnnotationHub)
ah = AnnotationHub()
cgi = ah[["AH5086"]]
```

```{r}
##We re-run the code from above
library(qvalue)
library(limma)
X = model.matrix(~pd$Status)
fit = lmFit(meth,X)
eb = eBayes(fit)
pvals = eb$p.value[,2]
qvals = qvalue(pvals)$qvalue
index = which(qvals<=0.05)

##Now we can see which CpGs are in islands
mean(gr[index]%over%cgi)
```

```{r}
islands=gr[index]%over%cgi
hypermethylated=fit$coef[index,2]>0
prop.table( table(islands,hypermethylated) )
```

### Q6

```{r}
library(bumphunter)
X = model.matrix(~pd$Status)
chr = as.character(seqnames(gr))
res = bumphunter(meth,X,chr=chr,pos=start(gr),cutoff=0.1)
```

```{r}
head(res$table)
```

```{r}
dmrs = res$table[ res$table$L>=3, ]
```

```{r}
dmrs = makeGRangesFromDataFrame(dmrs)
```

```{r}
##dmrs and cgis are defined above
map = distanceToNearest(dmrs,cgi)
d = mcols(map)$distance
mean(d==0)
```

### Q7

```{r}
map = distanceToNearest(dmrs,cgi)
d = mcols(map)$distance
mean(d>0 & d<=2000)
```

## Reading Raw 450K Array Data Assessment

### Q1

```{r}
BiocManager::install(c("minfi","IlluminaHumanMethylation450kmanifest","IlluminaHumanMethylation450kanno.ilmn12.hg19"))
```

```{r}
path <- "idats"
list.files(path)
```

```{r}
targets=read.csv(file.path(path,"targets.csv"),as.is=TRUE)
names(targets)
targets$Basename
```

```{r}
table(targets$Status)
```

### Q2

```{r}
targets$Basename = file.path(path,targets$Basename)
```

```{r}
library(minfi)
rgset = read.metharray(targets$Basename,verbose=TRUE)
rownames(targets)=sampleNames(rgset)
  pData(rgset)=targets
```

```{r}
dim(getRed(rgset))
dim(getGreen(rgset))
```

```{r}
mset = preprocessIllumina(rgset)
```

```{r}
mset = mapToGenome(mset)
```

```{r}
dim(getBeta(mset,type="Illumina")) ##the argument type="Illumina" gives us default procedure
head(granges(mset))
```

```{r}
i = which(seqnames(granges(mset))=="chr4" & start(granges(mset))==153807318)
j = which(rownames(pData(mset))=="5775041068_R04C01")
getBeta(mset,type="Illumina")[i,j]
```

### Q3

```{r}
library(bumphunter)
class(mset)
showMethods("bumphunter")
grset = ratioConvert(mset,what="beta",type="Illumina")
help("bumphunter,GenomicRatioSet-method")
```


```{r}
X=model.matrix(~pData(grset)$Status); res=bumphunter(grset,X,cutoff=0.1)
```

## CpGIslandShores assessment

```{r}
library(minfi)
library(IlluminaHumanMethylation450kmanifest)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
```

### Q1

```{r}
path <- "C:/Users/bioch/Downloads/tcgaMethylationSubset-master/tcgaMethylationSubset-master"
targets=read.delim(file.path(path,"targets.txt"),as.is=TRUE)
targets
```

```{r}
nrow(targets)
```

### Q2

```{r}
sum(targets$Tissue=="colon" & targets$Status=="normal")
##or look here
table(targets$Tissue,targets$Status)
```

### Q3

```{r}
index = which( targets$Status=="normal" & targets$Tissue%in%c("colon","breast") )
targets = targets[index,]
library(minfi)
dat = read.metharray.exp(base=path,targets = targets, verbose=TRUE)
class(dat)
## preprocess the data
dat = preprocessIllumina(dat)
class(dat)
## assign locations to each CpG
dat = mapToGenome(dat)
class(dat)
## precompute methylation values from U and M values
dat = ratioConvert(dat,type="Illumina")
class(dat)
library(rafalib)
mypar(1,1)
##extract methylation values
y = getBeta(dat)
shist(y)
```

```{r}
mds = cmdscale( dist(t(y)))
tissue = as.factor(pData(dat)$Tissue)
plot(mds,col=tissue)
```

```{r}
library(limma)
##create design matrix
tissue = as.factor(pData(dat)$Tissue)
X = model.matrix(~tissue)
##extract methylation values
y = getBeta(dat)
## obtain effect sizes and pvals with limma
fit = lmFit(y,X)
```

```{r}
index = which.max(abs( fit$coef[,2]))
seqnames(dat)[index]
start(dat)[index]
```

### Q4

```{r}
library(qvalue)
##create design matrix
tissue = as.factor(pData(dat)$Tissue)
X = model.matrix(~tissue)
##extract methylation values
y = getBeta(dat)
## obtain effect sizes and pvals with limma
fit = lmFit(y,X)
eb = eBayes(fit)
## obtain q-values
qvals = qvalue(eb$p.value[,2])$qvalue
```

```{r}
##When reporting such small p-values 
##there is no need to show exactly how small it is
## so we can say < 10^-6
index = which.max(abs( fit$coef[,2]))
qvals[index]
```

### Q5

```{r}
library(rafalib)
mypar(3,1)
index = which.max(abs( fit$coef[,2]))
gr=granges(dat)[index]+5000
index=which(granges(dat)%over%gr)
pos= start(dat)[index]

matplot(pos,y[index,],ylab="Methylation",col=as.numeric(tissue))

plot(pos, fit$coef[index,2],ylab="Effect Size")

plot(pos, -log10(qvals[index]) ,ylab="-log10 q-value")
```

### Q6

```{r}
library(rafalib)
mypar(3,1)
o = order(abs(fit$coef[,2]), decreasing = TRUE)[1:10]
for(i in o){
  index = i
  gr=granges(dat)[index]+5000
  index=which(granges(dat)%over%gr)
  pos= start(dat)[index]

  matplot(pos,y[index,,drop=FALSE],ylab="Methylation",col=as.numeric(tissue))

  plot(pos, fit$coef[index,2],ylab="Effect Size")

  plot(pos, -log10(qvals[index]) ,ylab="-log10 q-value")
}
```

### Q7

```{r}
index= which(seqnames(dat)=="chr15")
dat2 = dat[index,]
library(doParallel)
ncores = detectCores()
registerDoParallel(cores = ncores)
```

```{r}
##create design matrix
tissue = as.factor(pData(dat)$Tissue)
X = model.matrix(~tissue)
##extract methylation values
set.seed(1)
res = bumphunter(dat2,X,cutoff=0.1,B=100)
head(res$tab)
```

```{r}
sum(res$table$fwer < 0.05)
```

### Q8

```{r}
##fit and qvals were defined in a previous answer
index = which(qvals < 0.05 & abs(fit$coef[,2]) > 0.5 & seqnames(dat)=="chr15")
```

```{r}
tab = res$tab[ res$tab$L >= 3,]
tab = makeGRangesFromDataFrame(tab)
```

```{r}
mean(granges(dat[index,])%over% tab)
```

### Q9

```{r}
library(AnnotationHub)
cgi = AnnotationHub()[["AH5086"]]
```

```{r}
tab = res$tab[res$tab$fwer <= 0.05,]
tab = makeGRangesFromDataFrame(tab)
```

```{r}
map=distanceToNearest(tab,cgi)
d = mcols(map)$dist
mean( d<=2000 & d>0)
```

### Q10

```{r}
targets=read.delim(file.path (path,"targets.txt"),as.is=TRUE)
index = which( targets$Status=="normal" & targets$Tissue%in%c("colon","lung") )
targets = targets[index,]
```

```{r}
library(minfi)
dat = read.metharray.exp(base=path,targets = targets, verbose=TRUE)
```

```{r}
## preprocess the data
dat = preprocessIllumina(dat)
dat = mapToGenome(dat)
dat = ratioConvert(dat,type="Illumina")
```

```{r}
tissue=pData(dat)$Tissue
X = model.matrix(~tissue)
res = bumphunter(dat,X,cutoff=0.25)
nrow(res$tab)
```

```{r}
mean(res$tab$L==1)
```

### Q11

```{r}
load(file.path(path,"tcgaLungColonExpLM.rda"))
class(tcgaLungColonExpLM)
```

```{r}
print( annotation )  
```

```{r}
library(limma)
library(qvalue)
eb=eBayes(tcgaLungColonExpLM)
qvals=qvalue(eb$p.value[,2])$qvalue
```

```{r}
library(hgu133plus2.db)
map=select(hgu133plus2.db,keys=rownames(tcgaLungColonExpLM$coef),columns=c("ENTREZID"),keytype="PROBEID")

library(Homo.sapiens)
Genes=genes(Homo.sapiens)
Genes=resize(Genes,1) ## we want the tss

index1=match(as.character(mcols(Genes)$GENEID),map$ENTREZID)
index2 = match(map$PROBEID[index1],rownames(tcgaLungColonExpLM$coef))
M = tcgaLungColonExpLM$coef[index2,2]
```

```{r}
tab=makeGRangesFromDataFrame(res$tab,keep.extra.columns = TRUE)
map2=distanceToNearest(tab,Genes)
```

```{r}
index1=subjectHits(map2)
dist = mcols(map2)$dist


keep = dist< 10000 & tab$L>0
M[index1]
plot(tab$value,M[index1],main=paste("cor=",signif(cor(tab$value[keep],M[index1][keep],use="complete"),2)))


mypar(2,2)
for(i in c(0,1,2,3)){
  keep = dist< 10000 & tab$L>i
  plot(tab$value,M[index1],main=paste("cor=",signif(cor(tab$value,M[index1],use="complete"),2)))
}


library(rafalib)
mypar(2,2)
for(i in c(0,1,2,3)){
  keep = dist< 10000 & tab$L>i
  plot(tab$value[keep],M[index1][keep],main=paste("cor=",signif(cor(tab$value[keep],M[index1][keep],use="complete"),2)))
}
```

## Cell composition assessment

### Q1

```{r}
library(minfi)
grset=getGenomicRatioSetFromGEO("GSE32148")
```

```{r}
class(grset)
```

```{r}
##examine the outcome of this:
pData(grset)[1,]
## note age is here
pData(grset)[1,"characteristics_ch1.1"]
```

### Q2

```{r}
age=pData(grset)$characteristics_ch1.1
age=as.character(age)
age[grep("N/A",age)] = NA
age=as.numeric(gsub("age \\(y\\):\ ","",age))
```

```{r}
group = rep("normal",nrow(pData(grset)))
group[grepl("ulcerative",pData(grset)[,1])]="ulcerative"
group[grepl("Crohn",pData(grset)[,1])]="crohn"
group = factor(group,levels=c("normal","ulcerative","crohn"))
```

```{r}
keep = which(rowSums(is.na(getBeta(grset)))==0 &
               !seqnames(grset)%in%c("chrX","chrY"))
##create a new object 
grset2=grset[keep,]
```

```{r}
y = getBeta(grset2)
mds = cmdscale( dist( t(y)))
mypar(1,1)
col=ifelse(age>40,1,2)
plot(mds,pch=as.numeric(group),col=col)
legend("bottomleft",levels(group),pch=seq_along(group))
```

## Blocks assessment

### Q1

```{r}
library(minfi)
targets=read.delim(file.path (path,"targets.txt"),as.is=TRUE)
index = which( targets$Tissue=="colon")
targets = targets[index,]
dat = read.metharray.exp(base=path,targets = targets, verbose=TRUE)
```

```{r}
dat = preprocessIllumina(dat)
dat = mapToGenome(dat)
```

```{r}
cdat = cpgCollapse(dat)
```

```{r}
nrow(dat)
```

```{r}
nrow(cdat$obj)
```

### Q2

```{r}
head(granges(cdat$obj))
```

```{r}
prop.table(table(mcols( granges(cdat$obj))$type))
```

### Q3

```{r}
status = factor(pData(cdat$obj)$Status,
                level=c("normal","cancer"))
X=model.matrix(~status)
res = blockFinder(cdat$obj,X,cutoff=0.05)
```

```{r}
head(res$table)
```

```{r}
mean(res$table$value<0)
```

```{r}
tab=makeGRangesFromDataFrame(res$table)
index= granges(cdat$obj)%over% (tab[1]+10000)
pos=start(cdat$obj)[index]
col=as.numeric(status)
matplot(pos,getBeta(cdat$obj)[index,],col=col,pch=1,cex=0.5)
##and these are the estimated difference
plot(pos,res$fitted[index])
```

## Measuring methylation from sequencing

### Q1

```{r}
library("BSgenome.Hsapiens.UCSC.hg19")
chr22 = Hsapiens[["chr22"]]
```

```{r}
res=matchPattern("CCGG",chr22)
length(res)
```

### Q2

```{r}
size=diff(start(res))
hist(log10(size))
```

### Q3

```{r}
mean(size<=220 & size>=40)
```

### Q4

```{r}
sum( size[size<=220 & size>=40] )
```

### Q5

```{r}
path <- "C:/Users/bioch/Downloads/colonCancerWGBS-master"
targets = read.table(file.path(path,"targets.txt"), header = TRUE, sep = "\t")
targets
```

```{r}
library("bsseq")
cov.files = list.files(path=path,pattern="*chr22.cov",full.names=TRUE)  # coverage files
colonCancerWGBS =read.bismark(files=cov.files, rmZeroCov=TRUE, colData = DataFrame(targets))
###Note you might see a warning message here. You can ignore.
```

```{r}
colonCancerWGBS

# phenotypic information
pData(colonCancerWGBS)

# granges object
granges(colonCancerWGBS)

cov=getCoverage(colonCancerWGBS,type = "Cov")
m=getCoverage(colonCancerWGBS,type = "M")
```

```{r}
index=apply(cov>0,1,all)
mean(index)
```

### Q6

```{r}
tot = rowSums(cov)
##there are some very large values
hist(tot)
loc= start(colonCancerWGBS)
##plot by pieces
for(i in 1:11){
  index=1:100000+100000*i ##very ad-hoc
  plot(loc[index],tot[index],cex=.5,ylim=c(0,300))
}
```

### Q7

```{r}
cov=getCoverage(colonCancerWGBS,type = "Cov")
m=getCoverage(colonCancerWGBS,type = "M")
```

```{r}
gr = GRanges(seqnames="22",ranges=IRanges(start=43793678,end= 45022550))
index=granges(colonCancerWGBS)%over%gr
library(rafalib)
i=1
index2=which(index & cov[,i]>=5 & cov[,i]<=50)
x=start(colonCancerWGBS)[index2]
y=m[index2,i]/cov[index2,i]
w=sqrt(cov[index2,i])/7
plot(x,y,cex=w)
```

