---
title: "stats-gen-quiz3"
author: "Amy Gill"
date: "4/14/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Load the Montgomery and Pickrell eSet:

```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
```

What percentage of variation is explained by the 1st PC in the dataset if you:

1. Do no transformations?
2. log2(data+1) transform?
3. log2(data+1) transform and subtract row means?

**- 0.89 0.97 0.35**
- 0.97 0.97 0.35
- 0.97 0.97 0.97
- 0.35 0.35 0.35

### Answer 1

```{r}
edata_log <- log2(edata + 1)
edata_log_centered <- edata_log - rowMeans(edata_log)

svd1 <- svd(edata)
pcs_1 <- svd1$d^2/sum(svd1$d^2)
pcs_1[1]

svd2 <- svd(edata_log)
pcs_2 <- svd2$d^2/sum(svd2$d^2)
pcs_2[1]

svd3 <- svd(edata_log_centered)
pcs_3 <- svd3$d^2/sum(svd3$d^2)
pcs_3[1]
svd3$v[1,]
```

## Question 2

Use the Montgomery Pickrell eSet. Perform the log2(data + 1) transform and subtract row means from the samples. Set the seed to 333 and use k-means to cluster the samples into two clusters. Use `svd` to calculate the singular vectors. What is the correlation between the first singular vector and the sample clustering indicator?

- -0.52
- 0.33
**- 0.87** I think
- 0.84

### Answer 2

```{r}
set.seed(333)
k_clusters <- kmeans(t(edata_log_centered), centers = 2)
cor(k_clusters$cluster, svd3$v[,1])
```


## Question 3

Load the Bodymap data with the following command:

```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)
```

Fit a linear model relating the first gene's counts to the number of technical replicates, treating the number of replicates as a factor. Plot the data for this gene versus the covariate. Can you think of why this model might not fit well?

**- There is only one data point with a value of 6 so it is likely that the estimated value of that number of technical replicates is highly variable***?*
- The data are right skewed
- There may be different numbers of counts for different numbers of technical replicates
- The difference between 2 and 5 technical replicates is not the same as the difference between 5 and 6 technical replicates

### Answer 3

```{r}
library(broom)
gene1 <- edata[1,]
techreps <- pdata_bm$num.tech.reps
fit <- lm(gene1~factor(techreps))
tidy(fit)
summary(factor(techreps))
pdata_bm
```


## Question 4

Use the Bodymap dataset. Fit a linear model relating the first gene's counts to the age of the person and sex of the samples. What is the value and interpretation of the coefficient for age?

**- -23.91. This coefficient means that for each additional year of age, the count goes down by an average of 23.91 for a fixed sex.**
- -207.26. This coefficient means that for each additional year of age, the count goes down by an average of 207.26 for a fixed sex.
- -23.25. This coefficient means that there is an average decrease of 23.91 in the count variable per year within each gender.
- -22.26. This coefficient means that for each additional year of age, the count goes down by an average of 207.26 for a fixed sex.

### Answer 4

```{r}
fit <- lm(gene1 ~ pdata_bm$age + pdata_bm$gender)
lm(fit)
```


## Question 5

Use the Montgomery and Pickrell eSet:

```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp = montpick.eset
pdata=pData(mp)
edata=as.data.frame(exprs(mp))
fdata = fData(mp)
```

Perform the log2(data+1) transform. Then fit a regression model to each sample using population as the outcome. Do this using the `lm.fit` function (hint: don't forget the intercept). What is the dimension of the residual matrix, the effects matrix and the coefficients matrix?

- r 52580x129, e 52580x129, coef 2x52580
- r 52580x129, e 129x52580, coef 2x52580
- r 129x52580, e 129x52580, coef 129x52580
**- r 129x52580, e 129x52580, coef 2x52580**

### Answer 5

```{r}
mod <- model.matrix( ~ pdata$population)

fit <- lm.fit(mod, t(edata_log))
dim(fit$residuals)
dim(fit$effects)
dim(fit$coefficients)
```


## Question 6

Use the Montgomery Pickrell eSet. Perform the log2(data+1) transform. Then fit a regression model to each sample using population as the outcome. Do this using the `lm.fit` function. What is the effects matrix?

- the model residuals for all samples for each gene, with the values for each gene stored in the columns of the matrix
- the estimated fitted values for all samples for each gene, with the values for each gene stored in the rows of the matrix
- the model coefficients for all samples for each gene, with the values for each gene stored in the rows of the matrix
- the model coefficients for all samples in each gene, with the values for each gene stored in the columns of the matrix
**- the estimated fitted values for each gene, with the values for each gene stored in the columns of the matrix**

### Answer 6

```{r}
fit$effects[1:2,1:6]
```


## Question 7

Use the Bodymap data:

```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)
```

Fit many regression models to the expression data where `age` is the outcome variable using the `lmFit` function from the `limma` package (hint: you may have to subset the expression data to the samples without missing values of age to get the model to fit). What is the coefficient for age for the 1,000th gene? Make a plot of the data and fitted values for this gene. Does the model fit well?

**- -27.61. The model doesn't fit well since there are two large outlying values and the rest of the values are near zero**
- -27.61. The model doesn't fit well since there appears to be a non-linear trend in the data.
- 2469.87. The model doesn't fit well since there appears to be a non-linear trend in the data.
- -27.61. The model fits well since there seems to be a flat trend in the data.

### Answer 7

```{r}
library(limma)
edata_w_age <- edata[,-which(is.na(pdata_bm$age))]
mod <- model.matrix( ~ pdata_bm$age)
mod
dim(edata_w_age)
dim(mod)
fit <- lmFit(edata_w_age, mod)
fit$coefficients[1000,]
```

```{r}
edata_w_age[1000,]
```

```{r}
plot(edata_w_age[1000,], fitted(fit)[1000,])
```


## Question 8

Use the Bodymap data. Fit many regression modesls to the expression data where `age` is the outcome variable and `tissue.type` is an adjustment variable using the `lmFit` function from the `limma` package (hint: you may have to subset the expression data to the samples without missing values of age to get the model to fit). What is wrong with this model?


**- `tissue.type` has 18 levels but there are only 16 data points per gene, so this model can't fit a unique solution**
- this model doesn't fit well because most genes have high values for all of the counts, but gene 1000 has mostly low values
- this model doesn't fit well because there are a large number of outliers for the white blood cell tissue
- the model doesn't fit well since age should be treated as a factor variable

## Question 9

Why is it difficult to distinguish the study effect from the population effect in the Montgomery Pickrell dataset from ReCount?

**- The effects are difficult to distinguish because each study only measured one population**
- the study effects and population effects are difficult to distinguish because the population effect is not adjusted for study
- The study effects and population effects are difficult to distinguish because the study effects are stronger
- The study effects and population effects are not difficult to distinguish since they are the same effect

### Answer 9

```{r}
table(pdata$population, pdata$study)
```


## Question 10

Load the Bodymap dataset:

```{r}
library(Biobase)
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
edata = exprs(bm)
pdata_bm=pData(bm)
```

Set the seed using the command `set.seed(33353)` then estimate a single surrogate variable using the `sva` function after log2(data+1) transforming the expression data, removing rows with row means less than 1, and treating age as the outcome (hint: you may have to subset the exression data to the samples without missing values of age to get the model to fit). What is the correlation between the estimated surrogate for batch and age? Is the surrogate more highly correlated with `race` or `gender`?

- Corr 0.99, more highly correlated with race
- Corr 0.2, more highly correlated with race
- Corr 0.33, more highly correlated with gender
**- Corr 0.2, more highly correlated with gender***?*

### Answer 10

```{r}
library(sva)
set.seed(33353)
bm_log <- log2(edata+1)
bm_filtered <- bm_log[rowMeans(bm_log) > 1,-which(is.na(pdata_bm$age))]
pdata_bm_filtered <- pdata_bm[-which(is.na(pdata_bm$age)),]

dim(bm_filtered)
```





```{r}
mod <- model.matrix( ~ age, data=pdata_bm_filtered)
mod0 <- model.matrix( ~ 1, data = pdata_bm_filtered)

sva1 <- sva(bm_filtered, mod, mod0, n.sv=1)
```

```{r}
summary(lm(sva1$sv ~ pdata_bm_filtered$age))

cor(sva1$sv[,1], pdata_bm_filtered$age)
cor(sva1$sv[,1], as.numeric(pdata_bm_filtered$gender))
cor(sva1$sv[,1], as.numeric(pdata_bm_filtered$race))

```

