---
title: "Untitled"
author: "Amy Gill"
date: "7/26/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Started Exercises #1

Read in the file femaleMiceWeights.csv and report the exact name of the column containing the weights.

```{r}
dat <- read.csv("femaleMiceWeights.csv")
head(dat)    # inspect first 6 rows
names(dat)    # show column names
names(dat[2])    # name of second column
```

## Getting Started Exercises #2

The `[` and `]` symbols can be used to extract specific rows and specific columns of the table. What is the entry in the 12th row and second column?

```{r}
dat[12,2]
```

## Getting Started Exercises #3

You should have learned how to use the `$` character to extract a column from a table and return it as a vector. Use `$` to extract the weight column and report the weight of the mouse in the 11th row.

```{r}
dat$Bodyweight[11]
```

## Getting Started Exercises #4

The `length` function returns the number of elements in a vector. How many mice are included in our dataset?


```{r}
length(dat$Bodyweight)
```

## Getting Started Exercises #5

To create a vector with the numbers 3 to 7, we can use seq(3,7) or, because they are consecutive, 3:7. View the data and determine what rows are associated with the high fat or hf diet. Then use the mean function to compute the average weight of these mice.

```{r}
mean(dat$Bodyweight[13:24])
```

## Getting Started Exercises #6

```{r}
set.seed(1, sample.kind = "Rounding")
i <- sample( 13:24, 1)
dat$Bodyweight[i]
```

