---
title: "Untitled"
author: "Amy Gill"
date: "7/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r}
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extd\
ata/mice_pheno.csv"
filename <- basename(url)
download(url, destfile=filename)
dat <- read.csv(filename)
dat <- na.omit(dat)
```

## Question 1

Use **dplyr** to create a vector `x` with the body weight of all males on the control (chow) diet.
What is this population’s average?

```{r message=FALSE}
library(tidyverse)
x <- dat %>% filter(Sex == "M", Diet == "chow") %>% pull(Bodyweight)
mean(x)
```

## Question 2

Now use the rafalib package and use the `popsd` function to compute the population
standard deviation.

```{r}
popsd(x)
```

## Question 3

Set the seed at 1. Take a random sample X of size 25 from x. What is the sample average?

```{r}
set.seed(1, sample.kind = "Rounding")
X <- sample(x, 25)
mean(X)
```

## Question 4

Use dplyr to create a vector `y` with the body weight of all males on the high fat (hf) diet.
What is this population’s average?

```{r}
y <- dat %>% filter(Sex == "M", Diet == "hf") %>% pull(Bodyweight)
mean(y)
```

## Question 5

Now use the rafalib package and use the popsd function to compute the population
standard deviation.

```{r}
popsd(y)
```

## Question 6

Set the seed at 1. Take a random sample Y of size 25 from y. What is the sample average?

```{r}
set.seed(1, sample.kind = "Rounding")
Y <- sample(y, 25)
mean(Y)
```

## Question 7

What is the difference in absolute value between $\bar{y} - \bar{x}$ and $\bar{X}-\bar{Y}$?

```{r}
abs(mean(y)-mean(x))-abs(mean(X)-mean(Y))
```

## Question 8

Repeat the above for females. Make sure to set the seed to 1 before each sample call. What
is the difference in absolute value between $\bar{y} -\bar{x}$ and $\bar{X}-\bar{Y}$ ?

```{r}
x <- dat %>% filter(Sex == "F", Diet == "chow") %>% pull(Bodyweight)
set.seed(1, sample.kind = "Rounding")
X <- sample(x, 25)

y <- dat %>% filter(Sex == "F", Diet == "hf") %>% pull(Bodyweight)
set.seed(1, sample.kind = "Rounding")
Y <- sample(y, 25)

abs(mean(y)-mean(x))-abs(mean(X)-mean(Y))
```

## Question 9

For the females, our sample estimates were closer to the population difference than with
males. What is a possible explanation for this?

**A) The population variance of the females is smaller than that of the males; thus, the sample variable has less variability.**
B) Statistical estimates are more precise for females.
C) The sample size was larger for females.
D) The sample size was smaller for females.