---
title: "Statistics for Genomic Data Science - Week 2"
author: "Amy Gill"
date: "4/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r}
tropical = c("darkorange", "dodgerblue", "hotpink", "limegreen", "yellow")
palette(tropical)
par(pch = 19)

library(devtools)
library(Biobase)
```

```{r}
con <- url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/montpick_eset.RData")
load(file=con)
close(con)
mp <- montpick.eset
pdata <- pData(mp)
edata <- as.data.frame(exprs(mp))
fdata <- fData(mp)
```

```{r}
# making data easier to work with and visualize
edata <- edata[rowMeans(edata) > 100,]    # remove genes with mean expression < 100
edata <- log2(edata + 1)    # log scale to improve visualization ability
```

center the data before SVD/PCA - otherwise, the first PC will always be the means in a genomic experiment
(center == remove mean from all observation, in this case rowMeans)

then run SVD

```{r}
edata_centered <- edata - rowMeans(edata)    # center gene data - lets us find diffs between genes
svd1 <- svd(edata_centered)    # run SVD, generating matrices d, u, v
```

```{r}
dim(svd1$v)
dim(svd1$u)

# plot the singular values
plot(svd1$d, ylab = "Singular value", col = 2)
```

```{r}
# plot percent variance explained
plot(svd1$d^2/sum(svd1$d^2), ylab = "Percent Variance Explained", col=2)
```

Try to determine what the variable could be by coloring by study.

```{r}
par(mfrow=c(1,2))
plot(svd1$v[,1], col=2, ylab = "1st PC")
plot(svd1$v[,2], col=2, ylab = "2nd PC")
```

```{r}
par(mfrow = c(1,2))
plot(svd1$v[,1], svd1$v[,2], col=2, ylab = "2nd PC", xlab = "1st PC")
plot(svd1$v[,1], svd1$v[,2], ylab = "2nd PC", xlab = "1st PC",    # same plot colored by study
     col = as.numeric(pdata$study))

```

```{r}
boxplot(svd1$v[,1] ~ pdata$study, border = c(1,2))
points(svd1$v[,1] ~ jitter(as.numeric(pdata$study)), col = as.numeric(pdata$study))
```

Principal components

If you don't scale the data the same way, then the PCs and SVD right singular vectors ("principal components" casually) are not the same:

```{r}
pc1 <- prcomp(edata)
plot(pc1$rotation[,1], svd1$v[,1])
```

But if you scale them the same way, then they are:

```{r}
edata_centered_2 <- t(t(edata) - colMeans(edata))    # centered by column instead of by row
svd2 <- svd(edata_centered_2)
plot(pc1$rotation[,1], svd2$v[,1])
```

This is because the PCA involves centering by column instead of by row. 


Note that outliers strongly drive SVD. Here we add a single outlier row and show that the singular vectors don't match anymore:

```{r}
edata_outlier <- edata_centered
edata_outlier[6,] <- edata_centered[6,]*10000
svd3 <- svd(edata_outlier)
plot(svd1$v[,1], svd3$v[,1], xlab = "Without outlier", ylab = "With outlier")
```

If you plot the new singular vectors against the value of the outlier row, you see a almost perfect correlation, indicating that the SVD has been driven by the outlier.

```{r}
plot(svd3$v[,1], edata_outlier[6,], col=4)
```

Thus it is very important that all your features are on a common scale. If some gene is significantly overexpressed compared to others, then that gene can drive the apparent differences across samples and mask actual significant differences.